---
title: "Marginal Predictions for GLMMs with predglmm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Marginal Predictions for GLMMs with predglmm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Introduction

Standard predict methods for generalized linear mixed models return either
conditional predictions (using estimated random effects) or predictions with
random effects set to zero. Neither of these is the marginal mean
$E(Y_i \mid X_i)$, which averages over the random effects distribution:
$$
E(Y_i \mid X_i) = \int h(\eta_i + z_i^\top b) \, \varphi(b; 0, \Psi) \, db,
$$
where $h$ is the inverse link function, $\eta_i = x_i^\top \beta$ is the
linear predictor, $z_i$ is the random effects design vector for observation
$i$, and $\varphi(\cdot; 0, \Psi)$ is the density of the random effects.

When $h$ is nonlinear, $E[h(\eta + z^\top b)] \neq h(\eta)$ by Jensen's
inequality. The `predglmm` package computes the correct marginal mean.

For common link functions, the integral has a closed form:

- **Identity link** ($h(x) = x$): $E(Y_i \mid X_i) = \eta_i$.
- **Log link** ($h(x) = e^x$): $E(Y_i \mid X_i) = \exp(\eta_i + \sigma_i^2 / 2)$,
  where $\sigma_i^2 = z_i^\top \Psi \, z_i$.
- **Square-root link** ($h(x) = x^2$): $E(Y_i \mid X_i) = \eta_i^2 + \sigma_i^2$.

For other link functions, the package uses Gauss--Hermite quadrature.

## Example: Poisson GLMM with square-root link

We simulate data from a Poisson GLMM inspired by a behavioral experiment.
The design has four treatment groups, three experimental phases, and nine
time bins per phase. Subjects are observed once per bin, giving 27
observations per subject. The fixed effects allow separate mean functions
for each group--phase combination with polynomial time trends. The random
effects include correlated phase-specific intercepts per subject and
observation-level intercepts to capture overdispersion.

```{r simulate}
library(predglmm)
library(glmmTMB)
library(lme4)
library(ggplot2)

set.seed(42)

n_per_group <- 50
n_groups <- 4
n_subjects <- n_groups * n_per_group
n_phases <- 3
n_bins <- 9

# Design
subject <- factor(rep(1:n_subjects, each = n_phases * n_bins))
group <- factor(rep(rep(1:n_groups, each = n_per_group),
                    each = n_phases * n_bins))
phase <- factor(rep(rep(1:n_phases, each = n_bins), n_subjects))
bin <- rep(rep(1:n_bins, n_phases), n_subjects)

d <- data.frame(subject, group, phase, bin)

# Fixed effects on the sqrt scale:
# Group baselines
mu_group <- c(1.1, 1.4, 0.9, 1.0)
# Phase shifts
phase_shift <- c(0, 4.0, 0.9)
# Time trend (linear in log-bin)
time_slope <- c(-0.3, 0.5, -0.2)

d$eta <- mu_group[d$group] + phase_shift[d$phase] +
  time_slope[d$phase] * log(d$bin)

# Correlated random effects per subject (one per phase)
Psi <- matrix(c(
   0.59, -0.59, -0.33,
  -0.59,  1.75,  0.74,
  -0.33,  0.74,  0.92
), 3, 3)
L <- t(chol(Psi))
b <- matrix(rnorm(n_subjects * 3), nrow = 3)
b <- L %*% b  # 3 x n_subjects

# Map subject RE to observations
b_obs_re <- numeric(nrow(d))
for (i in 1:nrow(d)) {
  b_obs_re[i] <- b[d$phase[i], d$subject[i]]
}

# Observation-level RE
sigma_obs <- 0.87
e <- rnorm(nrow(d), sd = sigma_obs)

# Response (sqrt link: mu = eta^2)
d$eta_full <- d$eta + b_obs_re + e
d$y <- rpois(nrow(d), lambda = pmax(d$eta_full^2, 0))
d$obs <- factor(1:nrow(d))
```

### Fitting the model

We fit the model with `glmmTMB`. The fixed effects specify separate group
and phase intercepts with a polynomial trend in log-time within each
phase--group combination. The random effects include correlated
phase-specific intercepts per subject and observation-level intercepts.

```{r fit-glmmtmb, cache = TRUE}
fit_tmb <- glmmTMB(
  y ~ 0 + group * phase * log(bin) + (0 + phase | subject) + (1 | obs),
  family = poisson(link = "sqrt"),
  data = d
)
```

### Computing marginal predictions

The `pred_glmmtmb` function extracts the fixed effects, random effects
design, and variance components from the fitted model, then computes the
marginal mean.

```{r predictions}
d$marginal <- pred_glmmtmb(fit_tmb)
d$zero_re <- predict(fit_tmb, type = "response", re.form = NA)
d$conditional <- predict(fit_tmb, type = "response")
```

### Visualization

We compare the three prediction types, averaged over subjects within each
group--phase--bin combination.

```{r plot-comparison, fig.width = 8, fig.height = 5}
comp <- aggregate(
  cbind(marginal, zero_re, conditional, y) ~ group + phase + bin,
  data = d, FUN = mean
)

comp_long <- reshape(comp,
  varying = c("marginal", "zero_re", "conditional", "y"),
  v.names = "value",
  timevar = "type",
  times = c("Marginal", "Zero RE", "Conditional", "Observed"),
  direction = "long"
)
comp_long$type <- factor(comp_long$type,
  levels = c("Observed", "Marginal", "Zero RE", "Conditional"))

# Continuous time axis across phases
comp_long$time <- (as.numeric(comp_long$phase) - 1) * n_bins + comp_long$bin

ggplot(comp_long, aes(x = time, y = value,
                      color = type, linetype = type)) +
  geom_line(data = subset(comp_long, type != "Observed"), linewidth = 0.6) +
  geom_point(size = 1.5, data = subset(comp_long, type == "Observed")) +
  facet_wrap(~ group,
    labeller = labeller(group = function(x) paste("Group", x))) +
  scale_color_manual(values = c(
    "Observed" = "black",
    "Marginal" = "#2166ac",
    "Zero RE" = "#d6604d",
    "Conditional" = "#bababa"
  )) +
  scale_linetype_manual(values = c(
    "Observed" = "blank",
    "Marginal" = "solid",
    "Zero RE" = "dashed",
    "Conditional" = "dotted"
  )) +
  geom_vline(xintercept = c(9.5, 18.5), linetype = "solid",
             color = "gray70") +
  labs(x = "Time bin", y = "Rate", color = NULL, linetype = NULL) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

The marginal predictions (blue, solid) are consistently larger than the
zero-RE predictions (red, dashed). This gap reflects Jensen's inequality:
since the square-root link is convex, ignoring the random effects variance
leads to systematic underestimation. The conditional predictions
(gray, dotted) average out to the marginal curve because they incorporate
estimated random effects for each subject.

### Cross-validation with lme4

```{r fit-lme4, cache = TRUE}
fit_lme4 <- glmer(
  y ~ 0 + group * phase * log(bin) + (0 + phase | subject) + (1 | obs),
  family = poisson(link = "sqrt"),
  data = d
)
```

```{r compare-packages}
d$marginal_lme4 <- pred_glmer(fit_lme4)
summary(d$marginal - d$marginal_lme4)
```

## Nonlinear mixed models with nlme

For nonlinear mixed models fit with `nlme`, the random effects enter the
model function directly as parameters. There is no link function; the
mean is a nonlinear function $f(x_i, \beta + b_i)$. The marginal mean is
$$
E(Y_i \mid X_i) = \int f(x_i, \beta + b) \, \varphi(b; 0, \Psi) \, db.
$$

The `pred_nlme` function computes this by Monte Carlo integration.

```{r nlme-example}
library(nlme)

fit_nlme <- nlme(
  height ~ SSasymp(age, Asym, R0, lrc),
  fixed = Asym + R0 + lrc ~ 1,
  random = Asym ~ 1 | Seed,
  data = Loblolly
)

set.seed(1)
pred_marginal <- pred_nlme(fit_nlme, nsim = 5000)
pred_population <- as.numeric(predict(fit_nlme, level = 0))
```

Here the random effect is on `Asym`, which enters linearly. Jensen's
inequality does not apply, so the marginal and population predictions
coincide:

```{r nlme-linear}
summary(pred_marginal - pred_population)
```

When the random effect is on a nonlinear parameter, the two diverge:

```{r nlme-nonlinear}
fit_nlme2 <- nlme(
  height ~ SSasymp(age, Asym, R0, lrc),
  fixed = Asym + R0 + lrc ~ 1,
  random = lrc ~ 1 | Seed,
  data = Loblolly
)

set.seed(1)
pred_marginal2 <- pred_nlme(fit_nlme2, nsim = 5000)
pred_population2 <- as.numeric(predict(fit_nlme2, level = 0))

summary(pred_marginal2 - pred_population2)
```
